%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% References
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{thebibliography}{99.}%
\addcontentsline{toc}{chapter}{Bibliography} % Add to TOC

\bibitem{Brooks-1987} Brooks, F.P. (1987). No Silver Bullet: Essence and Accidents of Software Engineering, Computer, IEEE Computer Society Press, vol. 20, no. 4, pp. 10-19.

\bibitem{Brooks-1995} Frederick P. Brooks, Jr. The Mythical Man-Month. Addison-Wesley, Anniversary edition, 1995.

\bibitem{Burns-1978} J. M. Burns. Leadership. Harper and Row, New York, 1978.

\bibitem{Glass-2003} Robert L. Glass. Facts and Fallacies of Software Engineering. Pearson Education, Boston, MA, 2003.

\bibitem{Lehman-1980a} Meir M. Lehman (1980). ``Programs, Life Cycles, and Laws of Software Evolution''. Proc. IEEE, vol. 68, no. 9, pp. 1060-1076.

\bibitem{Lehman-1980b} Meir M. Lehman (1980). ``On Understanding Laws, Evolution, and Conservation in the Large-Program Life Cycle''. Journal of Systems and Software, vol. 1, pp. 213-221.

\bibitem{Lehman-1997} Meir M. Lehman, J. F. Ramil, P. D. Wernick, D. E. Perry, and W. M. Turski (1997). ``Metrics and laws of software evolution -- the nineties view''. Proc. 4th International Software Metrics Symposium (METRICS '97). pp. 20-32.

\bibitem{Yu-2013} Liguo Yu and Alok Mishra (Nov. 2013). ``An Empirical Study of Lehman's Law on Software Quality Evolution''. International Journal of Software and Informatics, vol. 7, no. 3, pp.469-481.


\end{thebibliography}
